---
title: "Microeconomics of insurance"
subtitle: "Chapter 2"
bibliography: micro-references.bib
link-citations: true
toc: true
toc-location: right
toc-depth: 3
toc-title: "Chapter contents"
---

## Decision theory

Before studying the demand for insurance, we need a language for describing **preferences under risk**: how does an individual compare a sure wealth level with a risky one, or choose between two risky wealth levels? We begin with these preferences and then introduce a mathematical representation that will allow us to study risk aversion and insurance decisions.

### Preferences over risky wealth

Fix a probability space $(\Omega,\mathcal F,\mathbb P)$. We consider an agent's preferences over random **total wealth** levels $X>0$ almost surely. Throughout the chapter, wealth takes values in $\mathbb R_+^*=(0,\infty)$. A payoff or a loss may be zero; the resulting total wealth must remain strictly positive.

The agent **weakly prefers** a wealth $Y$ to a wealth $X$ if they either strictly prefer $Y$ or are indifferent between the two. The word *weakly* includes indifference.

We assume that these preferences admit an **expected-utility representation**: there exists a function $u:(0,\infty)\to\mathbb R$ such that

$$
X\precsim_u Y
\quad\Longleftrightarrow\quad
\mathbb E[u(X)]\leq\mathbb E[u(Y)].
$$

The function $u$ is a **utility function of wealth**. It represents the agent's preferences by assigning to each random wealth its expected utility. The notation $X\precsim_u Y$ expresses weak preference for $Y$ over $X$, with the subscript indicating the representing utility. Strict preference and indifference are denoted by $\prec_u$ and $\sim_u$:

$$
\begin{aligned}
X\prec_u Y
&\quad\Longleftrightarrow\quad \mathbb E[u(X)]<\mathbb E[u(Y)]
&&\text{(strict preference)},\\
X\sim_u Y
&\quad\Longleftrightarrow\quad \mathbb E[u(X)]=\mathbb E[u(Y)]
&&\text{(indifference)}.
\end{aligned}
$$

This is the expected-utility framework of von Neumann and Morgenstern, whose book *Theory of Games and Economic Behavior* (1944) introduced the framework, with an axiomatic derivation of numerical utility added in the second edition [@vonNeumannMorgenstern1947]. The point is not that individuals literally write down a utility function before making a decision. Rather, the function provides a mathematical representation of their preferences under risk.

::: {.callout-note}
## Utility is not unique

If $\alpha>0$ and $\beta\in\mathbb R$, then $u$ and $\widetilde u=\alpha u+\beta$ represent exactly the same preferences. Any economically meaningful quantity derived from utility should therefore be invariant under **increasing affine transformations**. The affine restriction matters: an arbitrary increasing transformation need not preserve expected-utility comparisons.

:::

::: {.callout-important}
## Standing assumptions

The following conventions apply throughout the chapter unless stated otherwise.

**Utility functions.** Every utility function, including comparison utilities denoted by $v$, belongs to $C^2((0,\infty))$ and satisfies, for every $x>0$,

$$
u'(x)>0,
\qquad
u''(x)\leq 0.
$$

We shall state it explicitly when the stronger condition $u''<0$ is required. Since $u$ is increasing, the limit

$$
u(0):=\lim_{x\downarrow0}u(x)\in\mathbb R\cup\{-\infty\}
$$

exists. This is only boundary notation: zero wealth is excluded, and no smoothness at zero is assumed.

**Positive wealth.** Every random variable used as an argument of a utility function is strictly positive almost surely. In each insurance model, explicit strict inequalities on wealth, losses and premiums will ensure this. When differentiating expected utility, we shall give bounds that keep wealth uniformly away from zero.

**Integrability.** We restrict attention to random variables for which all the expectations under consideration are well defined and finite. The precise integrability conditions depend jointly on the random variable and on the utility function.

**Monotonicity.** *Increasing* and *decreasing* mean strictly monotone. We use *non-decreasing* and *non-increasing* when equality is allowed.

:::

The condition $u'>0$ implies that more wealth is preferred to less. The condition $u''\leq0$ says that the marginal utility of wealth is non-increasing: an additional euro matters no more (and typically less) at a higher wealth level. If $u''(x)=0$ for all $x$, then $u$ is affine and increasing (we often normalise it to $u(x)=x$), and the agent is said to be risk-neutral.

The geometric meaning of concavity is useful to keep in mind. For $x_0,x_1>0$ and $\theta\in[0,1]$,

$$
u\bigl(\theta x_0+(1-\theta)x_1\bigr)
\geq
\theta u(x_0)+(1-\theta)u(x_1).
$$

The graph of a concave utility function therefore lies above every chord joining two points of the graph.

```{r}
#| label: fig-concave-utility
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 4.6
#| fig-cap: "The utility of an average wealth lies above the corresponding average utility because of concavity."
#| fig-alt: "Increasing concave utility curve with a chord between two points and dotted guides showing x0, x1, the intermediate wealth and the corresponding utility levels."

x <- seq(0.15, 4.2, length.out = 500)
u <- sqrt(x)

x0 <- 0.55
x1 <- 3.60
theta <- 0.45

xm <- theta * x0 + (1 - theta) * x1

u0 <- sqrt(x0)
u1 <- sqrt(x1)
um <- sqrt(xm)

uchord <- theta * u0 + (1 - theta) * u1

ylim <- c(0, 2.15)

op <- par(
  mar = c(3.4, 6.5, 2.0, 1.5),
  xaxs = "i",
  yaxs = "i"
)

plot(
  x, u,
  type = "l",
  lwd = 2.5,
  col = "#6f97b5",
  xlab = "",
  ylab = "",
  axes = FALSE,
  ylim = ylim
)

# x-axis labels
axis(
  1,
  at = c(x0, xm, x1),
  labels = c(
    expression(x[0]),
    expression(theta*x[0] + (1-theta)*x[1]),
    expression(x[1])
  ),
  tick = FALSE,
  line = -0.4,
  cex.axis = 0.75,
  col.axis = "grey35"
)

# y-axis labels
axis(
  2,
  at = c(u0, uchord, um, u1),
  labels = c(
    expression(u(x[0])),
    expression(theta*u(x[0]) + (1-theta)*u(x[1])),
    expression(u(theta*x[0] + (1-theta)*x[1])),
    expression(u(x[1]))
  ),
  tick = FALSE,
  las = 1,
  line = -0.4,
  cex.axis = 0.70,
  col.axis = "grey35"
)

box(col = "grey60")

# chord
segments(
  x0, u0,
  x1, u1,
  lwd = 2,
  lty = 2,
  col = "grey55"
)

# vertical dotted guides
abline(
  v = c(x0, xm, x1),
  lty = 3,
  col = "grey75"
)

# horizontal dotted guides
segments(
  0, u0,
  x0, u0,
  lty = 3,
  col = "grey75"
)

segments(
  0, u1,
  x1, u1,
  lty = 3,
  col = "grey75"
)

segments(
  0, um,
  xm, um,
  lty = 3,
  col = "grey75"
)

segments(
  0, uchord,
  xm, uchord,
  lty = 3,
  col = "grey75"
)

# points defining the chord
points(
  c(x0, x1),
  c(u0, u1),
  pch = 16,
  col = "grey45"
)

# point on the utility curve
points(
  xm, um,
  pch = 16,
  col = "#6f97b5"
)

# corresponding point on the chord
points(
  xm, uchord,
  pch = 1,
  lwd = 1.5,
  col = "grey45"
)

par(op)

```

Concavity has an immediate interpretation under uncertainty.

::: {#thm-jensen}
## Jensen's inequality

Let $u$ be concave and let $X$ be an integrable random wealth with $X>0$ almost surely and $u(X)$ integrable. Then

$$
\mathbb E[u(X)]
\leq
u\bigl(\mathbb E[X]\bigr).
$$

If $u$ is strictly concave and $X$ is **non-degenerate**, meaning that $X$ is not almost surely constant, the inequality is strict.

:::

Under our concavity assumption, Jensen's inequality always gives

$$
X\precsim_u\mathbb E[X].
$$

The agent weakly prefers a sure wealth level to any risky wealth with the same mean. When $u$ is strictly concave, the above preference becomes strict for every non-degenerate random variable. The latter property is called **risk aversion** and we say that the agent is **risk-averse**. Here we use *risk-averse* for this strict preference.

A simple classroom example illustrates risk aversion. Fix an initial wealth $w>0$ and interpret all the following amounts as **additional payoffs**, so a zero payoff leaves wealth $w$, not zero. Compare a sure €1 with a fair coin toss paying €0 or €2. At such small stakes, one may feel almost indifferent. Compare next a sure €10,000 with a fair gamble paying €0 or €20,000: the sure amount may now be much more attractive. Finally, consider a sure €250,000 versus a fair gamble paying €0 or €1,000,000. Choosing the sure amount, which most people in the classroom do, means accepting only half the expected payoff to avoid the uncertainty.

The last example raises a more informative question: what sure amount would make the individual indifferent to the gamble? Most people cannot specify their utility function -- and, honestly, I cannot either --, but they can often give an intuitive answer to such a choice, or say whether they feel a proposed entry price makes a game worth playing. This motivates the concept of certainty equivalent and, for purchases of risky payoffs, that of indifference price.

### Certainty equivalents

The previous example suggests a natural way of attaching a monetary value to a risky prospect.

::: {#def-certainty-equivalent}
## Certainty equivalent

Let $X>0$ almost surely and assume that $u(X)$ is integrable. The **certainty equivalent** of $X$ for utility $u$ is the unique sure wealth level $\operatorname{ce}_u(X)>0$ satisfying

$$
u\bigl(\operatorname{ce}_u(X)\bigr)
=
\mathbb E[u(X)].
$$

Equivalently,

$$
\operatorname{ce}_u(X)
=
u^{-1}\!\left(\mathbb E[u(X)]\right).
$$

:::

The definition makes sense: continuity and strict monotonicity imply that $u((0,\infty))$ is an open interval. The finite expectation $\mathbb E[u(X)]$ lies strictly between its endpoints, since $u(X)$ does so almost surely. Thus the inverse is evaluated in its range and the certainty equivalent is strictly positive, even when $u(0)=-\infty$.

In French, the same object is called *équivalent certain* and often denoted $ec_u(X)$.

The certainty equivalent is a monetary valuation of **final wealth**: it is the sure wealth level that makes the agent indifferent to the risky wealth $X$.

::: {.callout-note}
## Certainty equivalents and indifference prices

When the random variable represents an **incremental payoff** rather than final wealth, it is often more natural to speak of an **indifference price**.

For a simple bounded-payoff setting, suppose $Y\geq0$ is bounded and $w>\operatorname{ess\,sup}Y>0$. At any price $0\leq P<w$, final wealth satisfies

$$
w+Y-P\geq w-P>0.
$$

The buyer's **indifference price** $P_u(w,Y)$ is the unique solution in $[0,\operatorname{ess\,sup}Y]$ of

$$
\mathbb E\!\left[u\bigl(w+Y-P_u(w,Y)\bigr)\right]=u(w).
$$

Indeed, on $[0,\operatorname{ess\,sup}Y]$ the left-hand side is continuous and strictly decreasing in the price. It is at least $u(w)$ at $P=0$ and at most $u(w)$ at $P=\operatorname{ess\,sup}Y$. The intermediate value theorem gives existence and strict monotonicity gives uniqueness. The inequality $w>\operatorname{ess\,sup}Y$ keeps final wealth at least $w-\operatorname{ess\,sup}Y>0$ throughout this argument.

Among prices $0\leq P<w$, the agent prefers buying below $P_u(w,Y)$, is indifferent at that price, and prefers not buying above it. Thus $P_u(w,Y)$ is the maximum acceptable price.

In general, $P_u(w,Y)$ depends on initial wealth: it values the payoff in the context of the individual's existing resources. By contrast, $\operatorname{ce}_u(Y)$ treats $Y$ itself as final wealth and is therefore not defined in our framework if $Y=0$ with positive probability.

For CARA utility $u(x)=-e^{-\gamma x}$ with $\gamma>0$ (see below), factorising the exponential in the indifference equation gives

$$
P_u(w,Y)=-\frac1\gamma\log\mathbb E[e^{-\gamma Y}].
$$

This price does not depend on $w$ as long as the positivity condition holds. The formula does not require evaluating utility at the payoff $Y$ itself.

:::

::: {#prp-ce-mean}
## Certainty equivalent and the mean

For $X>0$ almost surely with $X$ and $u(X)$ integrable,

$$
\operatorname{ce}_u(X)
\leq
\mathbb E[X].
$$

If $u$ is strictly concave and $X$ is non-degenerate, then the inequality is strict.

:::

::: {.proof}
By Jensen's inequality,

$$
\mathbb E[u(X)]
\leq
u(\mathbb E[X]).
$$

By definition,

$$
\mathbb E[u(X)]
=
u\bigl(\operatorname{ce}_u(X)\bigr).
$$

Since $u$ is increasing,

$$
\operatorname{ce}_u(X)
\leq
\mathbb E[X].
$$

Strict concavity and non-degeneracy give the strict version through the strict form of Jensen's inequality.

:::

The difference

$$
\Pi_u(X)
:=
\mathbb E[X]-\operatorname{ce}_u(X)
\geq0
$$

is called the **risk premium**. It is the discount, relative to the expected value, associated with bearing the risk.

::: {.callout-note}
## The St Petersburg paradox

The St Petersburg paradox illustrates why expected monetary value alone need not determine choices under risk. Nicolas Bernoulli posed the problem in his correspondence in 1713. Nowadays, the game would roughly proceed as follows: toss a fair coin; if it lands heads, the game ends and you receive €2. Otherwise, toss again: a head on the second toss pays €4. If the first two tosses are tails, continue, with €8 paid for a first head on the third toss, and so on. Each tail doubles the amount at stake, and the game ends at the first head.

Let $N$ be the number of tosses until the first head and let $Z_n=2^n$ be the payoff if the game ends on toss $n$. A first head on toss $n$ requires $n-1$ tails followed by one head, so $\mathbb P(N=n)=2^{-n}$. The game ends almost surely, but its payoff $Z_N=2^N$ has infinite expectation (this example deliberately departs from our finite-expectation convention):

$$
\mathbb E[Z_N]
=\sum_{n=1}^{\infty}Z_n\mathbb P(N=n)
=\sum_{n=1}^{\infty}2^n2^{-n}
=\sum_{n=1}^{\infty}1=+\infty.
$$

Every possible value of the stopping time contributes the same positive amount to the expectation. Yet an infinite expected payoff does not make every entry fee attractive. In our framework, an agent with initial wealth $w>0$ may pay a fee $0\leq P<w+2$: final wealth then satisfies $w+Z_N-P\geq w+2-P>0$. Under logarithmic utility, $\mathbb E[\log(w+Z_N-P)]$ is finite, because the logarithm of the payoff grows only linearly in $N$ and $\mathbb E[N]=2$.

Daniel Bernoulli's celebrated 1738 response was to value wealth through a concave utility function rather than to maximise expected money directly [@Bernoulli1954] -- the date 1954 corresponding to an English translation of the original Latin paper. His central example was logarithmic utility. The episode is historically important because it makes the distinction between **money** and the **utility of money** explicit, long before the axiomatic formulation of expected utility by von Neumann and Morgenstern.

:::

### Understanding the risk premium and measuring risk aversion

The certainty equivalent depends both on the distribution of the risk and on the agent's utility function. For a small risk, these two effects separate particularly cleanly.

Let

$$
X_\varepsilon=\mu+\varepsilon Z,
$$

where $\mu>0$, $Z$ is bounded, $|\varepsilon|\|Z\|_\infty<\mu$, and

$$
\mathbb E[Z]=0,
\qquad
\mathbb E[Z^2]=1.
$$

Thus $\mu$ is the mean wealth and

$$
\operatorname{Var}(X_\varepsilon)=\varepsilon^2.
$$

::: {#prp-small-risk}
## Certainty equivalent of a small centred risk

Let $\mu>0$, and let $Z$ be bounded with

$$
\mathbb E[Z]=0,
\qquad
\mathbb E[Z^2]=1.
$$

For $|\varepsilon|\|Z\|_\infty<\mu$,

$$
0<\mu-|\varepsilon|\|Z\|_\infty\leq\mu+\varepsilon Z\leq\mu+|\varepsilon|\|Z\|_\infty.
$$

Then, as $\varepsilon\to0$,

$$
\operatorname{ce}_u(\mu+\varepsilon Z)
=
\mu
+
\frac12\frac{u''(\mu)}{u'(\mu)}\,\varepsilon^2
+
o(\varepsilon^2).
$$

:::

::: {.proof}
Taylor's formula with integral remainder gives

$$
u(\mu+\varepsilon Z)
=u(\mu)+\varepsilon Z u'(\mu)
+\varepsilon^2 Z^2\int_0^1(1-t)u''(\mu+t\varepsilon Z)\,dt.
$$

Taking expectations and using $\mathbb E[Z]=0$, we get

$$
\mathbb E[u(\mu+\varepsilon Z)]
=u(\mu)+\varepsilon^2\mathbb E\!\left[
Z^2\int_0^1(1-t)u''(\mu+t\varepsilon Z)\,dt\right].
$$

For $|\varepsilon|\|Z\|_\infty\leq\mu/2$, all arguments of $u''$ lie in $[\mu/2,3\mu/2]$. Continuity of $u''$ and dominated convergence therefore give

$$
\mathbb E\!\left[Z^2\int_0^1(1-t)u''(\mu+t\varepsilon Z)\,dt\right]
\longrightarrow \frac12u''(\mu)\mathbb E[Z^2]=\frac12u''(\mu).
$$

Consequently,

$$
\mathbb E[u(\mu+\varepsilon Z)]
=u(\mu)+\frac12u''(\mu)\varepsilon^2+o(\varepsilon^2).
$$

By definition of the certainty equivalent,

$$
\operatorname{ce}_u(\mu+\varepsilon Z)
=
u^{-1}\!\left(
u(\mu)
+
\frac12u''(\mu)\varepsilon^2
+
o(\varepsilon^2)
\right).
$$

A first-order expansion of $u^{-1}$ at $u(\mu)$ therefore yields

$$
\operatorname{ce}_u(\mu+\varepsilon Z)
=
\mu
+
\frac{1}{u'(\mu)}
\left(
\frac12u''(\mu)\varepsilon^2
+
o(\varepsilon^2)
\right),
$$

which is the stated result.

:::

The expansion separates two ingredients. The amount of risk is represented by the variance $\varepsilon^2$, while the local attitude towards risk is captured by the curvature ratio $-u''(\mu)/u'(\mu)$ evaluated **at the mean $\mu$**.

This motivates the fundamental definition.

::: {#def-ara}
## Risk aversion coefficients

The **coefficient of absolute risk aversion** associated with a utility function $u$ is

$$
\gamma_u(x)
:=
-\frac{u''(x)}{u'(x)}.
$$

For $x>0$, the corresponding **coefficient of relative risk aversion** is

$$
\rho_u(x)
:=
x\,\gamma_u(x).
$$

:::

The coefficient is classically associated with the work of Arrow and Pratt [@Arrow1971; @Pratt1964].

Because $u$ is concave and $u'>0$, $\gamma_u(x)\geq0$. More importantly, $\gamma_u$ is unchanged when $u$ is replaced by $\alpha u+\beta$ with $\alpha>0$: it therefore depends on preferences, not on the arbitrary choice of a utility representation.

With this notation, @prp-small-risk becomes

$$
\operatorname{ce}_u(\mu+\varepsilon Z)
=
\mu
-
\frac12\gamma_u(\mu)\varepsilon^2
+
o(\varepsilon^2),
$$

or, equivalently,

$$
\Pi_u(\mu+\varepsilon Z)
=
\frac12\gamma_u(\mu)\varepsilon^2
+
o(\varepsilon^2).
$$

Dropping the remainder gives the **Arrow-Pratt approximation**:

$$
\Pi_u(X)\approx\frac12\gamma_u(\mathbb E[X])\operatorname{Var}(X).
$$

It is a local approximation for risks concentrated near their mean, not an exact formula for arbitrary risks.

The leading-order cost of a small risk is therefore half the product of a **risk term**, its variance, and a **preference term**, the coefficient of absolute risk aversion evaluated at the mean wealth $\mu$.

### CARA utility

A particularly tractable case is that of constant absolute risk aversion (*CARA*).

::: {#prp-cara-characterisation}
## Characterisation of CARA utility

Suppose

$$
\gamma_u(x)=\gamma
$$

for every $x>0$, where $\gamma\geq0$ is constant. Then:

- if $\gamma=0$, then $u$ is affine and the agent is risk-neutral;
- if $\gamma>0$, then, up to an increasing affine transformation,

$$
u(x)=-e^{-\gamma x}.
$$

Conversely, an increasing affine utility function has coefficient zero, and, for $\gamma>0$, $u(x)=-e^{-\gamma x}$ has constant absolute risk aversion equal to $\gamma$.

:::

::: {.proof}
The equation $\gamma_u(x)=\gamma$ is

$$
-\frac{u''(x)}{u'(x)}
=
\gamma,
$$

or equivalently

$$
\frac{d}{dx}\log u'(x)
=
-\gamma.
$$

Hence

$$
u'(x)
=
C e^{-\gamma x},
\qquad C>0.
$$

If $\gamma>0$, integration gives

$$
u(x)
=
A-B e^{-\gamma x},
\qquad B>0,
$$

which is equivalent under an increasing affine transformation to $-e^{-\gamma x}$. If $\gamma=0$, then $u'$ is constant and $u$ is affine.

The rest of the proof is straightforward verification.

:::

CARA is a strong assumption for individuals because willingness to bear a fixed monetary risk does not depend on wealth, provided the resulting wealth remains strictly positive. It is nevertheless extremely convenient. In particular, for

$$
u(x)=-e^{-\gamma x},
\qquad \gamma>0,
$$

we have

$$
\operatorname{ce}_u(X)
=
-\frac1\gamma
\log\mathbb E[e^{-\gamma X}],
$$

for every $X>0$ almost surely. Moreover, for any constant $c$ such that $X+c>0$ almost surely,

$$
\operatorname{ce}_u(X+c)
=
\operatorname{ce}_u(X)+c.
$$

This translation invariance is one reason CARA specifications are common in models of firms or delegated decision-makers when one wants to analyse profit risk without modelling wealth effects explicitly.

The following bounded example gives an exact formula while keeping wealth strictly positive.

::: {#prp-cara-binary}
## CARA utility and a symmetric two-outcome risk

Let $u(x)=-e^{-\gamma x}$ for $x>0$, with $\gamma>0$, and let

$$
X=\mu+\sigma Z,\qquad
\mathbb P(Z=1)=\mathbb P(Z=-1)=\frac12,
\qquad 0\leq\sigma<\mu.
$$

Then $X\geq\mu-\sigma>0$ and

$$
\operatorname{ce}_u(X)
=\mu-\frac1\gamma\log\cosh(\gamma\sigma),
\qquad
\Pi_u(X)=\frac1\gamma\log\cosh(\gamma\sigma).
$$

In particular, as $\sigma\to0$,

$$
\Pi_u(X)=\frac12\gamma\sigma^2+O(\sigma^4),
$$

in agreement with the Arrow-Pratt approximation.

:::

::: {.proof}
Since $Z$ takes the values $-1$ and $1$ with equal probabilities,

$$
\mathbb E[e^{-\gamma X}]
=e^{-\gamma\mu}\frac{e^{-\gamma\sigma}+e^{\gamma\sigma}}2
=e^{-\gamma\mu}\cosh(\gamma\sigma).
$$

Taking $-\gamma^{-1}$ times the logarithm gives the certainty equivalent. Subtracting it from $\mathbb E[X]=\mu$ gives the risk premium. Finally, $\log\cosh t=t^2/2+O(t^4)$ as $t\to0$ gives the expansion. For $\sigma>0$, $\log\cosh(\gamma\sigma)<\gamma\sigma$, so the certainty equivalent is greater than $\mu-\sigma>0$; for $\sigma=0$, it equals $\mu$.

:::

### DARA, IARA and CRRA utility

Absolute risk aversion need not be constant. An agent exhibits

- **DARA** (*decreasing absolute risk aversion*) if $\gamma_u$ is decreasing with wealth;
- **IARA** (*increasing absolute risk aversion*) if $\gamma_u$ is increasing with wealth.

For individual decision-making, DARA is often regarded as more plausible than CARA: as wealth increases, a fixed monetary gamble becomes less important relative to the agent's resources. Some empirical work finds absolute risk aversion decreasing with wealth, but the evidence is not uniform and estimated risk attitudes vary substantially across people and elicitation methods. DARA is therefore best viewed as an economically useful benchmark rather than as a universal empirical law; see, for example, @GuisoPaiella2008 and the rather different insurance-choice evidence in @CohenEinav2007.

A second classical family imposes constant **relative** rather than absolute risk aversion.

::: {#prp-crra-characterisation}
## Characterisation of CRRA utility

Suppose

$$
\gamma_u(x)
=
\frac{\rho}{x},
\qquad
\rho>0.
$$

Then, up to an increasing affine transformation,

$$
u(x)=
\begin{cases}
\dfrac{x^{1-\rho}}{1-\rho},
& \rho\neq1,\\[1.1ex]
\log x,
& \rho=1.
\end{cases}
$$

Conversely, these utility functions satisfy

$$
\rho_u(x)
=
x\gamma_u(x)
=
\rho,
$$

and therefore have constant relative risk aversion.

:::

::: {.proof}
The equation

$$
\gamma_u(x)
=
\frac{\rho}{x}
$$

gives

$$
-\frac{u''(x)}{u'(x)}
=
\frac{\rho}{x},
$$

hence

$$
\frac{d}{dx}\log u'(x)
=
-\frac{\rho}{x}.
$$

Integrating,

$$
u'(x)
=
C x^{-\rho},
\qquad C>0.
$$

A second integration gives

$$
u(x)
=
A+B\frac{x^{1-\rho}}{1-\rho}
$$

when $\rho\neq1$, and

$$
u(x)
=
A+B\log x
$$

when $\rho=1$, with $B>0$. Increasing affine transformations remove the irrelevant constants.

:::

An agent with CRRA utility exhibits decreasing absolute risk aversion because

$$
\gamma_u(x)=\frac{\rho}{x}
$$

decreases with wealth. It also imposes a very specific rate of decline, and hence a rigid link between risk attitudes at different wealth levels: doubling wealth halves absolute risk aversion. The logarithmic case $\rho=1$ is the specification associated with Daniel Bernoulli's classical discussion of the St Petersburg paradox.

### Comparing risk aversion across agents

People can often answer questions about certainty equivalents -- *what sure amount would make you indifferent to this lottery?* -- much more readily than they can specify or draw an entire utility function. It is therefore useful to have a way of comparing the risk aversion and the certainty equivalent of two agents without assigning cardinal meaning to the vertical scales of their utility functions.

Consider one agent represented by utility $u$ and another represented by utility $v$. The key mathematical idea is that one agent is more risk-averse than the other precisely when their utility can be obtained from the other's by applying an additional increasing concave transformation.

::: {#prp-risk-aversion-comparison}
## Comparison of risk aversion

Let $u$ and $v$ be utility functions satisfying our standing assumptions on $(0,\infty)$. Then the following statements are equivalent:

1. for every $x>0$,

   $$
   \gamma_u(x)
   \geq
   \gamma_v(x);
   $$

2. there exists an increasing concave function $g$ on $v((0,\infty))$ such that

   $$
   u=g\circ v.
   $$

:::

In words, the agent represented by $u$ is more risk-averse than the agent represented by $v$ exactly when $u$ is an increasing concave transformation of $v$.

::: {.proof}
Because $v$ is increasing, it is invertible on its range. Define

$$
g
:=
u\circ v^{-1}.
$$

Then $u=g\circ v$, and $g$ is increasing because both $u$ and $v^{-1}$ are increasing. Differentiating the identity

$$
g(v(x))
=
u(x)
$$

gives

$$
g'(v(x))
=
\frac{u'(x)}{v'(x)}
>
0.
$$

Differentiating once more,

$$
g''(v(x))
=
\frac{u'(x)}{[v'(x)]^2}
\left(
\gamma_v(x)-\gamma_u(x)
\right).
$$

The factor

$$
\frac{u'(x)}{[v'(x)]^2}
$$

is positive. Hence

$$
g''\leq0
\quad\Longleftrightarrow\quad
\gamma_u\geq\gamma_v.
$$

This proves both directions.

:::

The comparison has an immediate implication for certainty equivalents.

::: {#prp-ce-ordering}
## Ordering of certainty equivalents

Suppose

$$
\gamma_u(x)
\geq
\gamma_v(x)
$$

for every $x>0$. Then, for every $X>0$ almost surely with $u(X)$ and $v(X)$ integrable,

$$
\operatorname{ce}_u(X)
\leq
\operatorname{ce}_v(X).
$$

:::

Thus, when the agent represented by $u$ is more risk-averse than the agent represented by $v$, they assign to every risky prospect a certainty equivalent no higher than the other agent's.

::: {.proof}
By @prp-risk-aversion-comparison, there exists an increasing concave function $g$ such that

$$
u=g\circ v.
$$

Jensen's inequality gives

$$
\mathbb E[u(X)]
=
\mathbb E[g(v(X))]
\leq
g\bigl(\mathbb E[v(X)]\bigr).
$$

By definition of the certainty equivalent under $v$,

$$
\mathbb E[v(X)]
=
v\bigl(\operatorname{ce}_v(X)\bigr).
$$

Therefore

$$
\mathbb E[u(X)]
\leq
g\!\left(
v\bigl(\operatorname{ce}_v(X)\bigr)
\right)
=
u\bigl(\operatorname{ce}_v(X)\bigr).
$$

But

$$
\mathbb E[u(X)]
=
u\bigl(\operatorname{ce}_u(X)\bigr).
$$

Since $u$ is increasing,

$$
\operatorname{ce}_u(X)
\leq
\operatorname{ce}_v(X).
$$

:::

## Demand for insurance

We now study the demand for insurance. We begin with the maximum amount an individual is willing to pay to eliminate a random loss. This willingness to pay depends on initial wealth, preferences and the loss distribution; it can be studied without specifying how insurers set prices.

We then introduce a pricing rule: the premium equals $\lambda\geq1$ times the expected reimbursement (the **pure premium**). The excess over the pure premium is the safety loading. Under this rule, we first ask how much proportional coverage the individual purchases, and then allow the indemnity function to be chosen freely, subject to reimbursement lying between zero and the loss. The question becomes: which shape of insurance contract maximises expected utility?

### Full insurance against a binary loss

We start with a binary loss to make the calculations explicit. The underlying approach extends to more general loss distributions, under suitable integrability assumptions.

An individual has initial wealth $w$, including an asset worth $S$, with $0<S<w$. Thus losing the entire asset still leaves strictly positive wealth $w-S$. The asset is completely lost with probability $p\in(0,1)$ and remains intact with probability $1-p$. The **loss** is modelled by the random variable

$$
X=\begin{cases}
0,&\text{with probability }1-p,\\
S,&\text{with probability }p.
\end{cases}
$$

Without insurance, the final wealth is $w-X$. Full insurance reimburses the entire loss in exchange for an **insurance premium** $0\leq\pi<w$, leaving the sure wealth $w-\pi$. The insurance premium is the price of the policy; it is distinct from the **risk premium** $\Pi_u$, which measures the utility cost of uncertainty in monetary terms.

The individual weakly prefers purchasing full insurance precisely when

$$
u(w-\pi)\geq\mathbb E[u(w-X)]
=(1-p)u(w)+p\,u(w-S).
$$

The parameter restrictions are therefore $p\in(0,1)$, $0<S<w$ and $0\leq\pi<w$. They ensure that both uninsured wealth $w-X\geq w-S$ and insured wealth $w-\pi$ are strictly positive.

::: {#prp-maximum-insurance-premium}
## Maximum premium for full insurance

The maximum premium the individual is willing to pay is

$$
\pi_{\max}(p,S,w)
=w-u^{-1}\!\left((1-p)u(w)+p\,u(w-S)\right)
=w-\operatorname{ce}_u(w-X).
$$

It satisfies

$$
pS\leq\pi_{\max}(p,S,w)<S.
$$

Moreover, $\pi_{\max}(p,S,w)$ is increasing in $p$ and in $S$, holding the other variables fixed.

:::

::: {.proof}
Because $p\in(0,1)$ and $u$ is increasing,

$$
u(w-S)<(1-p)u(w)+p\,u(w-S)<u(w).
$$

Continuity and strict monotonicity of $u$ therefore give a unique $\pi_{\max}(p,S,w)\in(0,S)$ satisfying the indifference equation

$$
u\!\left(w-\pi_{\max}(p,S,w)\right)
=(1-p)u(w)+p\,u(w-S).
$$

Applying $u^{-1}$ gives the stated formula. Since expected utility under full insurance decreases with its price, the agent weakly prefers insurance exactly for admissible premiums $\pi\leq\pi_{\max}(p,S,w)$.

Jensen's inequality gives

$$
\operatorname{ce}_u(w-X)\leq\mathbb E[w-X]=w-pS,
$$

hence $\pi_{\max}(p,S,w)\geq pS$.

Finally, recalling that

$$
(u^{-1})'(y)=\frac{1}{u'(u^{-1}(y))},
$$

we differentiate the explicit formula to obtain

$$
\frac{\partial\pi_{\max}}{\partial p}(p,S,w)
=\frac{u(w)-u(w-S)}{u'\!\left(w-\pi_{\max}(p,S,w)\right)}>0,
$$

and

$$
\frac{\partial\pi_{\max}}{\partial S}(p,S,w)
=\frac{p\,u'(w-S)}{u'\!\left(w-\pi_{\max}(p,S,w)\right)}>0.
$$

:::

The lower bound $pS=\mathbb E[X]$ is the **pure premium**, or actuarially fair premium, for full insurance. Willingness to pay exceeds this amount by exactly the risk premium of uninsured wealth:

$$
\pi_{\max}(p,S,w)=\mathbb E[X]+\Pi_u(w-X).
$$

### How does willingness to pay depend on wealth?

The effects of $p$ and $S$ are intuitive: a more likely loss or a more valuable asset makes full insurance more attractive. Asking the same question about wealth often divides the classroom. One intuition is that a wealthier individual can more easily replace the asset and therefore has less need for insurance. Another is that paying a given premium is less burdensome at higher wealth. Neither intuition alone answers the question: both the loss and the premium must be evaluated through the same utility function.

There is no general direction for the wealth effect unless additional assumptions are made. The answer depends in fact on how absolute risk aversion changes with wealth. DARA, IARA and CARA give clear answers. If absolute risk aversion is not monotone, the same individual may have willingness to pay that increases over some wealth ranges and decreases over others.

::: {#prp-full-insurance-wealth}
## Wealth and the maximum insurance premium

For fixed $p\in(0,1)$ and $S>0$, consider $w>S$:

- if $\gamma_u$ is non-increasing, then $\pi_{\max}(p,S,w)$ is non-increasing in $w$;
- if $\gamma_u$ is non-decreasing, then $\pi_{\max}(p,S,w)$ is non-decreasing in $w$;
- if $\gamma_u$ is constant, then $\pi_{\max}(p,S,w)$ does not depend on $w$.

The first two conclusions are strict when $\gamma_u$ is decreasing or increasing, respectively.

:::

::: {.proof}
We apply Jensen's inequality to marginal utility. Differentiate the explicit expression

$$
\pi_{\max}(p,S,w)
=w-u^{-1}\!\left((1-p)u(w)+p\,u(w-S)\right).
$$

The derivative of the inverse function gives

$$
\begin{aligned}
\frac{\partial\pi_{\max}}{\partial w}(p,S,w)
&=1-\frac{(1-p)u'(w)+p\,u'(w-S)}
{u'\!\left(w-\pi_{\max}(p,S,w)\right)}\\
&=1-\frac{\mathbb E[u'(w-X)]}
{u'\!\left(w-\pi_{\max}(p,S,w)\right)}.
\end{aligned}
$$

To determine the sign, introduce

$$
\varphi=u'\circ u^{-1}
$$

on $u((0,\infty))$. Its derivative satisfies

$$
\varphi'(u(x))=\frac{u''(x)}{u'(x)}=-\gamma_u(x).
$$

Since $u$ is increasing, non-increasing absolute risk aversion is equivalent to a non-decreasing derivative $\varphi'$, hence to convexity of $\varphi$. Jensen's inequality then yields

$$
\begin{aligned}
\mathbb E[u'(w-X)]
&=\mathbb E[\varphi(u(w-X))]\\
&\geq\varphi\!\left(\mathbb E[u(w-X)]\right)\\
&=u'\!\left(w-\pi_{\max}(p,S,w)\right).
\end{aligned}
$$

Therefore $\frac{\partial\pi_{\max}}{\partial w}(p,S,w)\leq0$.

If $\gamma_u$ is non-decreasing, then $\varphi'$ is non-increasing and $\varphi$ is concave. Jensen's inequality reverses, giving $\frac{\partial\pi_{\max}}{\partial w}(p,S,w)\geq0$. If $\gamma_u$ is constant, then $\varphi$ is affine and the derivative is zero.

Strict monotonicity of $\gamma_u$ makes $\varphi$ strictly convex or strictly concave. Since $u(w-X)$ is non-degenerate, Jensen's inequality is then strict. This argument uses only the standing $C^2$ assumption on $u$.

:::

::: {.callout-note collapse="true"}
## Alternative proof: a wealth shift is a change of utility

Fix $h>0$ and define

$$
v(x)=u(x+h)
$$

for every $x>0$. Because $h>0$, $v$ is a utility function on $(0,\infty)$ satisfying the same standing assumptions. Take $w>S$, so $w-X\geq w-S>0$ and $w+h-X\geq w+h-S>0$. Evaluating wealth $w+h-X$ under $u$ is equivalent to evaluating wealth $w-X$ under $v$:

$$
\mathbb E[u(w+h-X)]=\mathbb E[v(w-X)].
$$

The corresponding certainty equivalents satisfy, for every $Y>0$ almost surely with $u(Y+h)$ integrable,

$$
\operatorname{ce}_v(Y)=\operatorname{ce}_u(Y+h)-h.
$$

Indeed,

$$
\begin{aligned}
u\!\left(\operatorname{ce}_v(Y)+h\right)
&=v\!\left(\operatorname{ce}_v(Y)\right)\\
&=\mathbb E[v(Y)]\\
&=\mathbb E[u(Y+h)]\\
&=u\!\left(\operatorname{ce}_u(Y+h)\right),
\end{aligned}
$$

and strict monotonicity of $u$ gives the identity. Consequently,

$$
\begin{aligned}
\pi_{\max}(p,S,w+h)
&=w+h-\operatorname{ce}_u(w+h-X)\\
&=w-\operatorname{ce}_v(w-X).
\end{aligned}
$$

Now $\gamma_v(x)=\gamma_u(x+h)$. If absolute risk aversion is non-increasing, then $\gamma_u(x)\geq\gamma_v(x)$: by @prp-risk-aversion-comparison, $u$ is an increasing concave transformation of $v$. The ordering of certainty equivalents in @prp-ce-ordering gives

$$
\operatorname{ce}_u(w-X)\leq\operatorname{ce}_v(w-X),
$$

and hence

$$
\pi_{\max}(p,S,w+h)
\leq w-\operatorname{ce}_u(w-X)
=\pi_{\max}(p,S,w).
$$

Under non-decreasing absolute risk aversion the comparison reverses; under constant absolute risk aversion it is an equality. Strict monotonicity gives a strictly concave comparison function in the appropriate direction, so Jensen's inequality in the certainty-equivalent comparison is strict for this non-degenerate loss.

:::

Thus, under decreasing absolute risk aversion, a wealthier individual is willing to pay less to eliminate the **same monetary loss**.

### Example: CRRA utility

Let $w>S$, $p\in(0,1)$ and $\rho>0$. For the CRRA utility introduced earlier,

$$
\pi_{\max}(p,S,w)=
\begin{cases}
w-\left((1-p)w^{1-\rho}+p(w-S)^{1-\rho}\right)^{1/(1-\rho)},
&\rho\neq1,\\[1.2ex]
w-w^{1-p}(w-S)^p,
&\rho=1.
\end{cases}
$$

Since $\gamma_u(w)=\rho/w$ is decreasing, $\pi_{\max}(p,S,w)$ is decreasing in $w$ for a fixed loss size $S$.

For logarithmic utility, the formula becomes

$$
\pi_{\max}(p,S,w)
=w\left[1-\left(1-\frac Sw\right)^p\right].
$$

For example, if $p=0.1$ and $S=20$, willingness to pay is approximately $2.21$ at $w=100$, and $2.10$ at $w=200$, compared with a pure premium of $2$.

::: {.callout-note}
## Possible exercises and extensions

In this model, an interesting exercise is to study $\pi_{\max}(p,S,w)$ as $w\to\infty$, keeping $p$ and $S$ fixed: find its limit and the first correction term. Compare this with the case where the asset value is a fixed fraction of wealth, $S=sw$ with $s\in(0,1)$.

Also, many variants of this initial model are possible. For example, suppose damage occurs with probability $p$, but the fraction of the asset lost is uniform on $(0,1)$. Write

$$
X=SBU,\qquad B\sim\mathcal B(p),
\qquad U\sim\operatorname{Unif}(0,1),
$$

with $B$ and $U$ independent, and retain $0<S<w$. Thus $w-X\geq w-S>0$. Then

$$
\mathbb E[u(w-X)]=(1-p)u(w)+p\int_0^1u(w-Sz)\,dz.
$$

Use this expression to find the maximum premium for full insurance, compare it with the pure premium $\mathbb E[X]=pS/2$, and study its dependence on wealth.

:::

### Proportional insurance and the safety loading

We now allow a general non-negative loss $X$ and write

$$
m=\mathbb E[X]>0.
$$

A **proportional insurance contract** reimburses $\alpha X$, where $\alpha\in[0,1]$ is the insured fraction. Its premium is

$$
P(\alpha)=\lambda\mathbb E[\alpha X]=\lambda\alpha m,
\qquad \lambda\geq1.
$$

This is a **pricing assumption**, often called the *expected value premium principle*: the price is proportional to the pure premium. It is not a definition of insurance pricing in general. Other rules can account differently for risk, expenses or capital requirements; we shall discuss premium calculation principles in the next chapter. Here we fix this rule to study the insured's response to prices.

The factor $\lambda$ is the premium multiplier. Equivalently, write $\lambda=1+\tau$, where $\tau\geq0$ is the **safety loading rate**: the amount charged above the pure premium is $\tau\alpha m$. Thus $\lambda=1$ (or $\tau=0$) means actuarially fair pricing.

For the rest of this chapter, assume $X$ is bounded and write $M=\operatorname{ess\,sup}X\in(0,\infty)$, the smallest almost-sure upper bound on the loss. Thus $0\leq X\leq M$ almost surely and $0<m\leq M$. We also suppose $u''<0$ on $(0,\infty)$.

For proportional insurance, impose the explicit condition

$$
\boxed{w>\max\{M,\lambda m\}}.
$$

This ensures positive wealth both without insurance and with full insurance. Every intermediate coverage level also gives positive wealth, as the following bound shows.

Final wealth and expected utility are

$$
Y_\alpha=w-X+\alpha X-\lambda\alpha m
=w-(1-\alpha)X-\lambda\alpha m,
\qquad
f(\alpha)=\mathbb E[u(Y_\alpha)].
$$

Indeed,

$$
Y_\alpha=(1-\alpha)(w-X)+\alpha(w-\lambda m),
\qquad
0<w-\max\{M,\lambda m\}\leq Y_\alpha\leq w.
$$

Thus $u$, $u'$ and $u''$ are bounded on the fixed interval containing every $Y_\alpha$, which justifies differentiation under expectations. The insured chooses $\alpha$ to maximise $f$ on $[0,1]$. All statements below about varying $\lambda$ are made subject to $1\leq\lambda<w/m$.

The analysis relies on a classical covariance inequality, recalled first.

::: {.callout-note}
## Chebyshev's inequality and rearrangement

For non-decreasing functions $a,b$ with $a(X)$, $b(X)$ and $a(X)b(X)$ integrable, **Chebyshev's covariance inequality** states that

$$
\mathbb E[a(X)b(X)]\geq\mathbb E[a(X)]\mathbb E[b(X)].
$$

Indeed, for an independent copy $X'$,

$$
2\operatorname{Cov}(a(X),b(X))
=\mathbb E\!\left[
\underbrace{(a(X)-a(X'))(b(X)-b(X'))}_{\geq0}
\right].
$$

The inequality is strict if $X$ is non-degenerate and $a,b$ are increasing.

In the discrete setting, the **rearrangement inequality** says that pairing two ordered sequences in the same order maximises their sum of products. Thus, for $a_1\leq\cdots\leq a_n$, $b_1\leq\cdots\leq b_n$ and any permutation $\sigma$,

$$
\sum_{i=1}^n a_i b_i\geq\sum_{i=1}^n a_i b_{\sigma(i)}.
$$

Averaging over all permutations gives the **Chebyshev sum inequality**,

$$
\frac1n\sum_{i=1}^n a_i b_i
\geq\left(\frac1n\sum_{i=1}^n a_i\right)
\left(\frac1n\sum_{i=1}^n b_i\right),
$$

which is the covariance inequality for a uniformly chosen index.

:::

::: {#prp-proportional-insurance}
## Optimal proportional coverage

Suppose $X$ is non-degenerate. The function $f$ is strictly concave, with

$$
f'(\alpha)=\mathbb E[(X-\lambda m)u'(Y_\alpha)],
\qquad
f''(\alpha)=\mathbb E[(X-\lambda m)^2u''(Y_\alpha)]<0.
$$

Define the threshold

$$
\lambda_{\mathrm{prop}}
=\frac{\mathbb E[Xu'(w-X)]}{m\,\mathbb E[u'(w-X)]}>1.
$$

Then the unique optimal insured fraction $\alpha^*$ satisfies:

- if $\lambda=1$, then $\alpha^*=1$;
- if $1<\lambda<\lambda_{\mathrm{prop}}$, then $\alpha^*\in(0,1)$ is the unique solution of $f'(\alpha^*)=0$;
- if $\lambda\geq\lambda_{\mathrm{prop}}$, then $\alpha^*=0$.

In particular, full proportional insurance is never optimal when $\lambda>1$.

:::

::: {.proof}
For each realisation of $X$, $Y_\alpha$ is affine in $\alpha$, so concavity of $u$ implies concavity of $f$. Differentiation gives the displayed formulas. Non-degeneracy of $X$ and $u''<0$ make the second derivative strictly negative.

At zero coverage,

$$
f'(0)
=\mathbb E[Xu'(w-X)]-\lambda m\,\mathbb E[u'(w-X)]
=m\,\mathbb E[u'(w-X)](\lambda_{\mathrm{prop}}-\lambda).
$$

Both $x\mapsto x$ and $x\mapsto u'(w-x)$ are increasing. The strict covariance inequality in the preceding note therefore gives

$$
\lambda_{\mathrm{prop}}
=1+\frac{\operatorname{Cov}(X,u'(w-X))}
{m\,\mathbb E[u'(w-X)]}>1.
$$

At full coverage, final wealth is the constant $w-\lambda m$, hence

$$
f'(1)=(1-\lambda)m\,u'(w-\lambda m).
$$

This is zero when $\lambda=1$ and strictly negative when $\lambda>1$. Since $f'$ is decreasing, its endpoint signs give all three cases.

:::

The last statement of @prp-proportional-insurance is known as **Mossin's theorem** [@Mossin1968].

The covariance term expresses the value of insurance: reimbursement arrives in states where losses are high and marginal utility of wealth is high. A safety loading reduces this benefit by charging more than the expected reimbursement.

At full insurance, however, marginal utility is the same in every state. A first-order argument in $\varepsilon>0$ explains the incentive to reduce coverage: moving from $1$ to $1-\varepsilon$ saves $\lambda\varepsilon m$ in premium and introduces a retained loss of mean $\varepsilon m$. More precisely,

$$
f(1-\varepsilon)-f(1)
=\varepsilon(\lambda-1)m\,u'(w-\lambda m)+O(\varepsilon^2).
$$

For $\lambda>1$, the gain in expected wealth is first order, while the utility cost of the small risk around sure wealth is only second order. This is why paying more than one euro per expected euro reimbursed cannot justify eliminating the very last part of the risk.

```{r}
#| label: fig-proportional-demand
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4.2
#| fig-cap: "Expected utility under proportional insurance: fair pricing gives full coverage, a moderate loading gives partial coverage, and a sufficiently high loading gives no coverage. Each curve is shifted by its value at zero coverage. Illustration with logarithmic utility, w = 100, and a loss of 40 with probability 0.2."
#| fig-alt: "Three strictly concave expected utility curves. Their maxima occur respectively at full coverage, partial coverage, and zero coverage."

alpha <- seq(0, 1, length.out = 501)
w <- 100
S <- 40
p <- 0.2
m <- p * S
loads <- c(1, 1.2, 1.6)
titles <- c("Fair premium", "Partial coverage", "No coverage")
op <- par(mfrow = c(1, 3), mar = c(4.3, 5.2, 2.4, 1.0), mgp = c(3.2, 0.8, 0))
for (j in seq_along(loads)) {
  lam <- loads[j]
  objective <- function(a) {
    (1-p) * log(w-lam*a*m) + p * log(w-S+a*S-lam*a*m)
  }
  values <- vapply(alpha, objective, numeric(1)) - objective(0)
  candidates <- c(0, optimize(objective, c(0, 1), maximum = TRUE)$maximum, 1)
  best <- candidates[which.max(vapply(candidates, objective, numeric(1)))]
  plot(alpha, values, type = "l", lwd = 2.5, col = "#6f97b5",
       xlab = expression(alpha), ylab = expression(f(alpha)-f(0)),
       main = titles[j], cex.main = 0.95,
       ylim = range(values) + c(-0.08, 0.30) * diff(range(values)))
  abline(h = 0, col = "grey75", lty = 3)
  abline(v = best, col = "grey55", lty = 2)
  points(best, objective(best)-objective(0), pch = 16, col = "#6f97b5")
  legend("topleft", legend = bquote(lambda == .(lam)),
         bty = "o", bg = "white", box.col = "grey65", inset = 0.03, cex = 0.9)
}
par(op)
```

### Choosing the shape of insurance: deductibles

Proportional insurance is only one possible contract. For the rest of the chapter, we strengthen the wealth condition to the simple sufficient inequality

$$
\boxed{w>M+\lambda m}.
$$

This allows us to compare all the indemnities below without imposing an additional solvency constraint on each contract. We now allow any measurable **indemnity function** $I$ satisfying

$$
0\leq I(x)\leq x,\qquad x\geq0.
$$

Thus the insurer never pays a negative amount and never reimburses more than the loss. These constraints alone permit a wide variety of shapes, including non-monotone indemnities. The next figure shows both familiar contracts and less conventional admissible examples.

```{r}
#| label: fig-indemnity-functions
#| echo: false
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 6.8
#| fig-cap: "Examples of admissible indemnities. Every curve lies between zero and the dashed full-insurance line I(x) = x. The last two examples satisfy the model's constraints even though reimbursement need not increase with the loss. All amounts are in arbitrary monetary units."
#| fig-alt: "Six indemnity curves: proportional insurance, a deductible, capped insurance, a capped deductible, an oscillating indemnity and a contract paying only for an intermediate band of losses. Each lies between zero and the loss."

loss_grid <- seq(0, 100, length.out = 1001)
indemnities <- list(
  function(x) 0.6 * x,
  function(x) pmax(x - 30, 0),
  function(x) pmin(x, 40),
  function(x) pmin(pmax(x - 20, 0), 35),
  function(x) x * (0.5 + 0.35 * sin(pi * x / 25)),
  function(x) ifelse(x >= 25 & x <= 65, 0.7 * x, 0)
)
contract_titles <- c("Proportional", "Deductible", "Capped",
                     "Capped deductible", "Oscillating", "Band coverage")
op <- par(mfrow = c(2, 3), mar = c(3.8, 4.1, 2.4, 0.8),
          mgp = c(2.3, 0.7, 0))
for (j in seq_along(indemnities)) {
  payment <- indemnities[[j]](loss_grid)
  stopifnot(all(payment >= 0), all(payment <= loss_grid))
  plot(loss_grid, payment, type = "n", xlim = c(0, 100), ylim = c(0, 100),
       xlab = "Loss x", ylab = "Indemnity I(x)",
       main = contract_titles[j], cex.main = 0.95)
  polygon(c(0, 100, 100), c(0, 0, 100), col = "#eef2f5", border = NA)
  abline(a = 0, b = 1, col = "grey60", lty = 2)
  if (j < 6) {
    lines(loss_grid, payment, col = "#6f97b5", lwd = 2.5)
  } else {
    # Draw the jumps explicitly rather than joining them by sloping lines.
    segments(c(0, 25, 65), c(0, 17.5, 0),
             c(25, 65, 100), c(0, 45.5, 0), col = "#6f97b5", lwd = 2.5)
    points(c(25, 65), c(17.5, 45.5), pch = 16, col = "#6f97b5")
    points(c(25, 65), c(0, 0), pch = 21, bg = "white", col = "#6f97b5")
  }
  box(col = "grey65")
}
par(op)
```

We keep the expected value premium principle introduced above, with the same multiplier $\lambda$ for every contract:

$$
P(I)=\lambda\mathbb E[I(X)].
$$

The insured seeks to maximise

$$
F(I)=\mathbb E\!\left[u\bigl(w-X+I(X)-\lambda\mathbb E[I(X)]\bigr)\right].
$$

For every such contract, $0\leq\mathbb E[I(X)]\leq m$, and hence

$$
0<w-M-\lambda m
\leq w-X+I(X)-\lambda\mathbb E[I(X)]\leq w.
$$

All final wealth levels therefore lie in the same positive interval $[w-M-\lambda m,w]$. The condition $w>M+\lambda m$ is sufficient, not necessary for each individual contract.

Here $F$ is a **functional**: its input is a function $I$, rather than a real number. The admissible indemnities form a convex subset of an infinite-dimensional function space.

Optimisation over functions often uses the **calculus of variations**. A classical source of the subject is the brachistochrone problem, posed by Johann Bernoulli in 1696: among curves connecting two points, find the one along which a particle descends under gravity in the shortest time. Euler and Lagrange subsequently developed systematic methods for such problems. Here the unknown function is an indemnity rather than a trajectory. In what follows, we need only the elementary variational idea of moving from one function towards another and differentiating along that direction.

A **deductible contract** with deductible $D\geq0$ has indemnity

$$
I_D(x)=(x-D)_+,
\qquad (z)_+=\max(z,0).
$$

The individual pays losses up to $D$, and the insurer pays the excess. Final wealth is therefore

$$
Y_D=w-X+(X-D)_+-\lambda\mathbb E[(X-D)_+].
$$

In particular, it equals $w-X-P(I_D)$ when $X\leq D$, and $w-D-P(I_D)$ when $X>D$.

Our strategy has two steps. Given any contract $I$, we first find a deductible contract $I_D$ with the **same premium**. We then show that this deductible gives at least as much expected utility. This will reduce the choice of an entire indemnity function to the choice of the single number $D$.

::: {#prp-matching-deductible}
## Matching the pure premium with a deductible

For every admissible indemnity $I$, there exists $D\in[0,M]$ such that

$$
\mathbb E[(X-D)_+]=\mathbb E[I(X)].
$$

If $\mathbb E[I(X)]>0$, then this deductible is unique and belongs to $[0,M)$. If $\mathbb E[I(X)]=0$, then the contract pays nothing almost surely; any $D\geq M$ gives the same zero indemnity, and we may choose $D=M$.

:::

::: {.proof}
The function

$$
H(D)=\mathbb E[(X-D)_+]
$$

is continuous: the inequality $|(x-D)_+-(x-E)_+|\leq|D-E|$ gives $|H(D)-H(E)|\leq|D-E|$. Moreover,

$$
H(0)=m,\qquad H(M)=0.
$$

Since $0\leq\mathbb E[I(X)]\leq m$, existence follows from the intermediate value theorem.

For uniqueness, recall that $M$ is the smallest almost-sure upper bound on $X$. If $0\leq D_1<D_2<M$, then $\mathbb P(X>D_2)>0$, and on this event the two indemnities differ by $D_2-D_1$. Hence $H(D_1)>H(D_2)$. Also $H(D)>0$ for $D<M$, so a positive target expectation is attained at exactly one deductible. Finally, a non-negative indemnity with zero expectation vanishes almost surely, as does $(X-D)_+$ for every $D\geq M$.

:::

::: {#thm-optimal-deductible}
## Optimality of deductible insurance

Let $I$ be any admissible indemnity and choose $D\in[0,M]$ such that

$$
\mathbb E[I_D(X)]=\mathbb E[I(X)].
$$

Then the two contracts have the same premium and

$$
F(I_D)\geq F(I).
$$

The inequality is strict if $\mathbb P(I_D(X)\neq I(X))>0$. Consequently, maximising expected utility over all admissible indemnities reduces to choosing a deductible $D\in[0,M]$.

:::

::: {.proof}
Write

$$
P=\lambda\mathbb E[I(X)]=\lambda\mathbb E[I_D(X)],
\qquad
\Delta(x)=I(x)-I_D(x).
$$

Interpolate between the two contracts:

$$
J_\theta=I_D+\theta(I-I_D),\qquad \theta\in[0,1].
$$

Every $J_\theta$ is admissible and has premium $P$, since $\mathbb E[\Delta(X)]=0$. Define

$$
g(\theta)=F(J_\theta)
=\mathbb E[u(w-X+J_\theta(X)-P)].
$$

For each realised loss, final wealth is affine in $\theta$ and lies in $[w-M-\lambda m,w]$. Thus $g$ is concave; boundedness of $u'$ on this interval justifies differentiation under the expectation, giving

$$
g'(0)=\mathbb E[u'(w-X+(X-D)_+-P)\Delta(X)].
$$

Split the expectation according to whether the loss exceeds the deductible:

$$
\begin{aligned}
g'(0)
&=\mathbb E\!\left[u'(w-X-P)\Delta(X)\mathbf 1_{\{X\leq D\}}\right]\\
&\quad+u'(w-D-P)\mathbb E\!\left[\Delta(X)\mathbf 1_{\{X>D\}}\right].
\end{aligned}
$$

Using $\mathbf 1_{\{X>D\}}=1-\mathbf 1_{\{X\leq D\}}$ and $\mathbb E[\Delta(X)]=0$,

$$
\mathbb E\!\left[\Delta(X)\mathbf 1_{\{X>D\}}\right]
=-\mathbb E\!\left[\Delta(X)\mathbf 1_{\{X\leq D\}}\right].
$$

On $\{X\leq D\}$, the deductible pays nothing, so $\Delta(X)=I(X)\geq0$. Since $u'$ is decreasing and $w-X-P\geq w-D-P$ on this event, we obtain

$$
\begin{aligned}
g'(0)
&=\mathbb E\!\left[
\underbrace{\bigl(u'(w-X-P)-u'(w-D-P)\bigr)
\mathbf 1_{\{X\leq D\}}}_{\leq0}
\underbrace{I(X)}_{\geq0}
\right]\\
&\leq0.
\end{aligned}
$$

Concavity now gives

$$
F(I)=g(1)\leq g(0)+g'(0)\leq g(0)=F(I_D).
$$

If $\Delta(X)$ is non-zero with positive probability, strict concavity of $u$ makes $g$ strictly concave. Its tangent inequality at zero is then strict at one, proving strict preference.

Finally, $D\mapsto F(I_D)$ is continuous on $[0,M]$: the indemnities and premiums vary continuously with $D$, and final wealth stays in $[w-M-\lambda m,w]$, where $u$ is bounded. Dominated convergence proves continuity, and compactness then gives a maximiser. The comparison above proves that this maximum is also optimal among all admissible indemnities.

:::

This result is due to Arrow [@Arrow1963; @Arrow1971].

For a fixed expected reimbursement, insurance is most valuable in states where the insured would otherwise be poorest. A deductible directs reimbursement towards large losses and equalises final wealth across all losses exceeding the deductible. Covering small losses instead uses part of the same premium to transfer money into states where marginal utility is lower.

### Choosing the optimal deductible

The remaining problem is a standard one-dimensional maximisation:

$$
\max_{D\in[0,M]}\psi(D),
\qquad
\psi(D)=F(I_D)
=\mathbb E\!\left[
 u\!\left(w-X+(X-D)_+-\lambda\mathbb E[(X-D)_+]\right)
\right].
$$

Continuity on the compact interval $[0,M]$ guarantees that an optimal deductible exists. Once the loss distribution and utility function are specified, it can be found analytically or numerically by the usual methods for a function of one variable, comparing candidate maximisers in $(0,M)$ with the endpoints $0$ and $M$. If the loss distribution has atoms, one should allow for points at which $\psi$ is not differentiable, using one-sided derivatives where appropriate.

## References

::: {#refs}
:::
